---
title:
author:
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
#description: "This is a minimal example of using the bookdown package to write a book. The output format for this example is bookdown::gitbook."
github-repo: childpovertyactionlab/wic-participation-reports
favicon: "images/favicon.ico"
---

# LA7 WIC Participation Report {-}

## LA7 Participation Summary {-}

```{r, include=FALSE}
rm(list=ls(all=TRUE))
###############################################################################################
theme_cpal <- function(base_size = 12, base_family = "sans") {
  colors <- deframe(ggthemes::ggthemes_data[["fivethirtyeight"]])
  (theme_foundation(base_size = base_size, base_family = base_family)
    + theme(
      line = element_line(colour = "#b4b4b4"),
      rect = element_rect(fill = "#ffffff",
                          linetype = 1, colour = NA),
      text = element_text(family = "Roboto", face = "bold", colour = "#6c6c6c"),
      axis.title = element_text(),
      axis.title.x = element_text(vjust = 2),
      axis.title.y = element_text(vjust = 2),
      axis.text = element_text(color = "#b4b4b4"),
      axis.ticks = element_blank(),
      #axis.ticks.length = unit(6, "pt"),
      axis.line = element_line(color = "#b4b4b4", size = 1.5, linetype = "solid"),
      legend.background = element_rect(),
      legend.position = "none",
      legend.direction = "horizontal",
      legend.box = "horizontal",
      panel.grid.major = element_line(colour = "#e1e1e1"),
      panel.grid.minor = element_blank(),
      plot.title = element_text(hjust = 0, size = rel(1.5), face = "bold"),
      plot.margin = unit(c(1, 3, 1, 1), "lines"),
      panel.border = element_rect(size=1, fill = NA),
      strip.background = element_rect()
    ))
}

CPAL.colors = c("#008097", "#ec008c", "#eaca2d", "#b4b4b4", "#9bd9e3", "#fdddd7", "#55107E")
CPAL.colors.Rev = c("#fdddd7", "#9bd9e3", "#b4b4b4", "#eaca2d", "#ec008c", "#008097")
CPAL.colors.Rev.Lim = c("#eaca2d", "#ec008c", "#008097")
###############################################################################################

library(data.table)
library(tidyverse)
library(rio)
library(ggthemes)
library(directlabels)
library(leaflet)
library(sf)
library(plotly)
library(tigris)
library(reactable)

WIC_Participation <- import("data/WIC_ParticipationReports.csv") %>%
  mutate(Fiscal_Month = factor(Month,
                               levels = c(10:12, 1:9),
                               labels = c(month.abb[10:12], month.abb[1:9])))

WIC_LA7 <- WIC_Participation %>%
  filter(Project == 7 & Year >= 2020)

la7participation <- import("data/LA7_ParticipationReport.csv") %>%
  janitor::clean_names() %>%
  mutate(ClinicNo = site)

dallas_sf <- counties(state = "TX") %>%
  filter(GEOID == 48113)

st_layers("E:/CPAL Dropbox/Basic Needs/04_Projects/WIC/WIC Eligibility/GIS/WIC_EligibilityEstimates.gpkg")

wic_eligible_sf <- st_read("E:/CPAL Dropbox/Basic Needs/02_Data/WIC Eligibility Estimates/WIC_Eligible_2018.shp") %>%
  st_transform(crs = 4269)

```

Between `r format(min(WIC_LA7$Date), "%B %Y")` and `r format(max(WIC_LA7$Date), "%B %Y")`, WIC participation in Local Agency 7 (Dallas County) changed from `r format(WIC_LA7$Total_Participants[which.min(WIC_LA7$Date)], big.mark = ",")` to `r format(WIC_LA7$Total_Participants[which.max(WIC_LA7$Date)], big.mark = ",")`, with a peak of `r format(max(WIC_LA7$Total_Participants), big.mark = ",")` participants in `r format(WIC_LA7$Date[which.max(WIC_LA7$Total_Participants)], "%B %Y")` a decline of `r format(round((WIC_LA7$Total_Participants[which.max(WIC_LA7$Date)]-WIC_LA7$Total_Participants[which.min(WIC_LA7$Date)])/WIC_LA7$Total_Participants[which.min(WIC_LA7$Date)]*100, digits = 2))`%.

This year has seen a continuation of the downward trend in participation seen over the past five years, with participation levels reaching their five year low on `r format(WIC_LA7$Date[which.min(WIC_LA7$Total_Participants)], "%B %Y")`.

```{r, fig.height=5, fig.width=7, fig.align='center', echo=FALSE, message=FALSE, warning=FALSE}
WICmonth <- WIC_Participation %>%
  filter(Project == 7) %>%
  filter(Year >= 2019) %>%
  ggplot(aes(x=Date, 
             y=Total_Participants, 
             )) +
  geom_line(color = "#008097", size = 1.5) +
  labs(title = "Local Agency 7 Participation by Month",
       subtitle = "",
       x = "",
       y = "") +
  scale_y_continuous(labels = scales::comma,
                     breaks = scales::breaks_width(2500),
                     limits = c(57500, 67500)) +
  scale_x_date(expand = expansion(mult = c(0.05, 0.1)),
               date_breaks = "1 year", date_minor_breaks = "6 months", date_labels = "%Y") +
  theme_cpal() +
  theme(legend.background = element_rect(),
      legend.position = "bottom",
      legend.direction = "horizontal",
      legend.box = "horizontal")

ggplotly(WICmonth, tooltip = c("Date", "Total_Participants")) %>%
  layout(legend = list(
    orientation = "h",
    y = -0.1,
    x = 0.2))
```

## Current Participation by Clinic {-}

```{r, include=FALSE}
setwd("E:/CPAL Dropbox/") #Michael Desktop Directory
#setwd("C:/Users/micha/CPAL Dropbox/") #Michael Laptop Directory

clinics <- st_read("Basic Needs/02_Data/WIC Clinics/WIC Clinics.gdb", layer = "WIC_Clinics") %>%
  st_transform(crs = 4326) %>%
  left_join(la7participation) %>%
  select(-(x17_oct:x19_dec)) %>%
  mutate(x21_apr = as.numeric(str_remove(x21_apr, ",")),
         x21_may = as.numeric(str_remove(x21_may, ",")))

sumOld <- sum(clinics$x20_sep, na.rm = TRUE)
sumNew <- sum(clinics$x21_sep, na.rm = TRUE)

#sumNew <-  as.data.frame(clinics) %>%
#  select (-Shape) %>%
#  .[ , ncol(.), drop = FALSE] %>%
#  sum(., na.rm = TRUE)

sumRecent <- sum(clinics$x21_sep, na.rm = TRUE)

clinics <- clinics %>%
  mutate(perpart = round((x21_sep/sumNew)*100, digits = 2),
         perchange = round((x21_sep-x20_sep)/sumOld*100, digits = 2),
         permonth = round((x21_sep-x21_aug)/sumOld*100, digits = 2))
```


```{r, fig.height=5, fig.width=7, echo=FALSE, message=FALSE, warning=FALSE}
WIClocal <- clinics %>%
  pivot_longer(cols = starts_with("x2"), names_to = "Date", values_to = "Participation") %>%
  mutate(Date = str_remove(Date, "x"),
         Date = lubridate::ym(Date)) %>%
  filter(Date == max(Date)) %>%
  drop_na(Participation) %>%
  ggplot(aes(x = reorder(site_name, Participation), 
             y = Participation)) +
  coord_flip() +
  geom_bar(stat = "identity", fill = "#008097", color = "#008097") +
  labs(title = "September 2021 Participation by Agency",
       subtitle = "Local Agency 7",
       x = "",
       y = "") +
  scale_y_continuous(labels = scales::comma) +
  theme_cpal()

ggplotly(WIClocal, tooltip = "Participation")
```

## Clinic Map {-}
```{r, fig.height=7, fig.width=7, echo=FALSE, message=FALSE, warning=FALSE}
cpal_style <- "https://api.mapbox.com/styles/v1/owencpal/ckecb71jp22ct19qc1id28jku/tiles/256/{z}/{x}/{y}@2x?access_token=pk.eyJ1Ijoib3dlbmNwYWwiLCJhIjoiY2tlYnR3emdxMGNhZzMwb2EzZWR4ajloNCJ9.P7Mujz8F3Rssq5-Q6dcvMw"

map_attr <- "© <a href='https://www.mapbox.com/map-feedback/'>Mapbox</a> Basemap © <a href='https://childpovertyactionlab.org/'>Child Poverty Action Lab</a>"

pal <- colorNumeric(
  palette = "BuPu",
  domain = wic_eligible_sf$WIC_ELIGI2)#,
#  n=7)

leaflet() %>%
  setView(lng = -96.7970, lat = 32.7767, zoom = 10) %>%
  addTiles(urlTemplate = cpal_style, attribution = map_attr) %>%
  addPolygons(data = wic_eligible_sf,
              fillOpacity = 0.8,
              fillColor = ~pal(WIC_ELIGI2),
              color = "#FFFFFF",
              weight = 1) %>%
  addPolygons(data = dallas_sf,
              fillOpacity = 0,
              weight = 5,
              color = "#008097") %>%
  addCircleMarkers(data = clinics,
                   radius = ~perpart*2,
                   stroke = FALSE,
                   fillOpacity = 0.5,
                   fillColor = "#008097",
                   popup = paste0("<b>", "Clinic: ", "</b>", clinics$site_name, "<br>",
                                 "<b>", "Total Participation: ", "</b>", clinics$mar_21, "<br>",
                                 "<b>", "Percent Participation: ", "</b>", clinics$perpart, "%", "<br>",
                                 "<b>", "YtY Change: ", "</b>", clinics$perchange, "%"))

```

```{r, echo=FALSE, results='asis'}
clinics %>%
  st_drop_geometry(.) %>%
  select(site_name, x21_sep, perpart, perchange, permonth) %>%
  mutate(perpart = paste0(perpart, "%"),
         perchange = paste0(perchange, "%"),
         permonth = paste0(permonth, "%")) %>%
  drop_na(x21_sep) %>%
  reactable(., columns = list(
    site_name = colDef(
      name = "Clinic"),
    x21_sep = colDef(
      name = "September 2021 Participation"),
    perpart = colDef(
      name = "Overall Participation"),
    perchange = colDef(
      name = "Percent Change (Year)"),
    permonth = colDef(
      name = "Percent Change (Month)")),
    theme = reactableTheme(
      headerStyle = list(
        "&:hover[aria-sort]" = list(background = "hsl(0, 0%, 96%)"),
        "&[aria-sort='ascending'], &[aria-sort='descending']" = list(background = "hsl(0, 0%, 96%)"),
        borderColor = "#555"
        )
      )
    )
```
## LA7 Participation Over Time {-}

Local Agency 7 experienced a brief period between January 2020 and July 2020 where participation change increased relative to a year prior. This has since trended downwards and participation is once more declining relative to a year ago, albeit at smaller levels than previously seen.

```{r, fig.height=5, fig.width=7, fig.align='center', echo=FALSE, message=FALSE, warning=FALSE}
WICyear <- WIC_Participation %>%
  filter(Project == 7) %>%
  mutate(Participants_Diff = Total_Participants - lag(Total_Participants, order_by = Date, n = 12),
         highlight = ifelse(Participants_Diff >= 0, "Gain", "Loss")) %>%
  ggplot(aes(x = as.Date(factor(Date)), 
             y = Participants_Diff,
             fill = highlight)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("#008097", "#ec008c")) +
  labs(title = "Year to Year WIC Participation Change",
       subtitle = "Local Agency 7",
       x = "",
       y = "Participation Change") + #color = "Year"
  scale_y_continuous(labels = scales::comma) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  theme_cpal()

ggplotly(WICyear, tooltip = c("Participants_Diff"))
```

## WIC Participation Across Texas {-}

Many local agencies in geographies similar to Dallas experienced brief increases in WIC participation during the first half of 2020. Local Agency 1, 17, and 41 are the only agencies who have maintained an increased number of families within the program in March 2021 relative to a year prior. All other local agencies have return to pre-pandemic levels of participation.

```{r, fig.height=5, fig.width=7, fig.align='center', echo=FALSE, message=FALSE, warning=FALSE}
WICurban <- WIC_Participation %>%
  mutate(highlight=ifelse(Project==7, "Dallas", 
                          ifelse(Project %in% c(17, 26, 48), "Houston",
                                 ifelse(Project == 1, "Austin", 
                                        ifelse(Project == 41, "San Antonio", 
                                               ifelse(Project == 54, "Ft. Worth", 
                                                      ifelse(Project == 0, "Texas", "Other"))))))) %>%
  filter(highlight != "Other") %>%
  group_by(Date, highlight) %>%
  mutate(Total_Participants = sum(Total_Participants)) %>%
  ggplot(aes(x=Date, y=Total_Participants, color = highlight, size = highlight)) +
  geom_line() +
  scale_color_manual(values = c("#008097", "#e98816", "#eaca2d", "#ea8b98", "#b4b4b4", "#9bd9e3")) +
  scale_size_manual(values = c(0.5, 1, 0.5, 0.5, 0.5, 0.5, 0.5, 0.5)) +
  labs(title = "Monthly WIC Participation in Urban Texas Counties",
       subtitle = "Between October 2015 through August 2021",
       x = "",
       y = "Average Participation") +
  scale_y_continuous(labels = scales::comma) +
  scale_x_date(expand = expansion(mult = c(0.05, 0.1)),
               date_breaks = "1 year", date_labels = "%Y") +
  guides(size = FALSE, color = FALSE) +
  theme_cpal() +
  theme(legend.background = element_rect(),
      legend.position = "bottom",
      legend.direction = "horizontal",
      legend.box = "horizontal")

ggplotly(WICurban, tooltip = c("highlight", "Total_Participants")) %>%
  layout(legend = list(
    orientation = "h",
    y = -0.1,
    x = 0))
```


